---
title: "epistack benchmarking"
author: "Guillaume Devailly"
package: epistack
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{epistack benchmarking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Generating benchmark datasets

To compare {epistack} with other methods, we will use an artificially generated
dataset. We will generate artificial datasets of various regon length using 
a function defined bellow (there is no need to understand how this function
works, it's simply generate a dumy object that can be input to 
`plotEpistack()`). 

```{r fakemat, message=FALSE}
library(GenomicRanges)
library(epistack)

epidataset <- function(n) {
  # fist we generate a fake epigenetic signal matrix
  mat = matrix(nrow = n, ncol = 50)
  for(i in seq_len(nrow(mat))) {
    mat[i, ] = runif(50) +
      c(sort(abs(rnorm(50)))[1:25], rev(sort(abs(rnorm(50)))[1:25]))*i/(n/4)
  }
  
  # we then embeded the matrix in a GRanges object
  testdat <- GRanges(
    rep("chr1", n),
    IRanges(rep(1, n), rep(2, n)),
    mcols = mat
  )
  testdat$expr <- exp(seq(from = 0, to = 3, length.out = n)) 
  testdat <- testdat[seq(length(testdat), 1),]
  testdat <- addBins(testdat, nbins = 5)
  testdat
}
```


The function is working as intended:
```{r testfunc}
epidataset(100)[, 49:52]
epidataset(10000)[, 49:52]
```


The datasets can then be plotted with epistack:

```{r plot1, fig.small=TRUE}

plotEpistack(
  epidataset(10000), 
  zlim = c(0, 3), ylim = c(0, 4), tints = "dodgerblue",
  patterns = "mcols", titles = "Mark level", x_labels = c("", "middle", ""), 
  metric_col = "expr", metric_title = "Expression"
)

```

We can now generate a benchmark dataset containing GRanges objects of
various lengths. 

```{r benchdata}
sizes <- c(
  1000, 2000, 5000,
  10000, 20000, 50000,
  100000, 200000
)

benchdata <- lapply(sizes, epidataset)

```

# Benchmarking `epistack()`

```{r benchmark1, message=FALSE, fig.keep = 'last'}

runtimes <- sapply(benchdata, function(x) system.time(
  plotEpistack(
    x, 
    zlim = c(0, 3), ylim = c(0, 4), tints = "dodgerblue",
    patterns = "mcols", titles = "Mark level", x_labels = c("", "middle", ""), 
    metric_col = "expr", metric_title = "Expression"
  )
)["elapsed"])

runtimes2 <- sapply(benchdata, function(x) system.time(
  plotEpistack(
    x, 
    zlim = c(0, 3), ylim = c(0, 4), tints = "dodgerblue",
    patterns = "mcols", titles = "Mark level", x_labels = c("", "middle", ""), 
    metric_col = "expr", metric_title = "Expression"
  )
)["elapsed"])

library(ggplot2)

data.frame(size = sizes, time = runtimes) |>
  ggplot(aes(x = size, y = time)) +
    geom_line() +
    geom_point(shape = 3) +
    theme_bw(base_size = 14) +
    labs(x = "Number of regions", y = "Time (s)",
         title = "plotEpistack() and EnrichedHeatmap() benchmark") +
    coord_cartesian(ylim = c(0, 15))

```

# Comparison with EnrichedHeatmap

First we transform the benchmark dataset into the format needed by 
EnrichedHeatmap:

```{r EnrichedHeatmap1}
library(EnrichedHeatmap)
# we first make a dummy normalizedMatrix object to later steal its attributes
load(
  system.file("extdata", "chr21_test_data.RData", package = "EnrichedHeatmap")
)
ehmat <- normalizeToMatrix(
  meth, 
  resize(cgi, width = 1, fix = "center"),
  value_column = "meth", mean_mode = "absolute",
  extend = 2500, w = 100, smooth = TRUE
)

# we then apply the attributes to the signal matrices in benchdata
ehbenchdata <- lapply(
  benchdata,
  function(x) {
    mat <- as.matrix(mcols(x)[, 1:50])
    mdim <- dim(mat)
    mdimnames <- dimnames(mat)
    mostattributes(mat) <- attributes(ehmat)
    dim(mat) <- mdim
    dimnames(mat) <- mdimnames
    mat
  }
)
```

We can now call EnrichedHeatmap:
```{r EnrichedHeatmap2}
col_fun = circlize::colorRamp2(c(0, 1.5, 3), c("white", "dodgerblue", "black"))
EnrichedHeatmap(
    ehbenchdata[[1]], col = col_fun,
    name= "Mark level",  column_title = "Mark level")

```

With some fiddeling, it is possible to make EnrichedHeatmap outputs
equivalent to the plotEpistack's outputs.

```{r EnrichedHeatmap3}
ht_list <- Heatmap(
  log10(benchdata[[1]]$expr+1), 
  col = c("white", "orange", "black"), name = "log10(expr+1)", 
  show_row_names = FALSE, width = unit(10, "mm"),
  top_annotation = HeatmapAnnotation(
    summary = anno_summary(
      gp = gpar(fill = 2:6),
      outline = FALSE, axis_param = list(side = "right")
    )
  )
) +
  Heatmap(
    paste("bin", benchdata[[1]]$bin),
    col = structure(2:6, names = paste("bin", 1:5)), 
    name = "bins",
    show_row_names = FALSE, width = unit(3, "mm")
  ) + 
  EnrichedHeatmap(
    ehbenchdata[[1]], col = col_fun,
    name= "Mark level",  column_title = "Mark level",
    top_annotation = HeatmapAnnotation(
      lines = anno_enriched(gp = gpar(col = 2:6))
    )
  )

draw(
  ht_list,
  split = paste("bin", benchdata[[1]]$bin), cluster_rows = FALSE
)

```

We can now run a benchmark. Note that there is currently 
[a bug](https://github.com/jokergoo/EnrichedHeatmap/issues/58) in 
EnrichedHeatmap preventing us to generate the Heatmaps with more than 100.000
regions unless we shut off rasterization.


```{r benchmark2, message=FALSE, fig.keep = 'last'}

runtimes2 <- sapply(1:8, function(i) system.time({
  
  ht_list <- Heatmap(
    log10(benchdata[[i]]$expr+1), 
    col = c("white", "orange", "black"), name = "log10(expr+1)", 
    show_row_names = FALSE, width = unit(10, "mm"),
    top_annotation = HeatmapAnnotation(
      summary = anno_summary(
        gp = gpar(fill = 2:6),
        outline = FALSE, axis_param = list(side = "right")
      )
    )
  ) +
    Heatmap(
      paste("bin", benchdata[[i]]$bin),
      col = structure(2:6, names = paste("bin", 1:5)), 
      name = "bins",
      show_row_names = FALSE, width = unit(3, "mm")
    ) + 
    EnrichedHeatmap(
      ehbenchdata[[i]], col = col_fun,
      name= "Mark level",  column_title = "Mark level",
      top_annotation = HeatmapAnnotation(
        lines = anno_enriched(gp = gpar(col = 2:6))
      )
    )
  
  draw(
    ht_list,
    split = paste("bin", benchdata[[i]]$bin), cluster_rows = FALSE,
    use_raster = FALSE
  )
  
})["elapsed"])


data.frame(
  size = c(sizes, sizes),
  time = c(runtimes, runtimes2),
  method = c(rep("epistack", 8), rep("EnrichedHeatmap", 8))) |>
  ggplot(aes(x = size, y = time, color = method)) +
    geom_line() +
    geom_point(shape = 3) +
    theme_bw(base_size = 14) +
    labs(x = "Number of regions", y = "Time (s)",
         title = "plotEpistack() benchmark") +
    coord_cartesian(ylim = c(0, 60))
```

EnrichedHeatmp is less fast, but can still plot up to hundred thousands
of regions!

# Making plotEpistack outputs similar to EnrichedHeatmap outputs

In the previous benchmark, we used the amazing flexibility of EnrichedHeatmap
to make its outputs similiar to epistack outputs. It is alsa possible to do the
opposti: make epistack output more similar to the EnrichedHeatmap outputs,
as illustrated bellow:

```{r enrichedepistack, fig.height=6}
testdata <- benchdata[[1]]
testdata$bins <- NULL

layout(matrix(1:3, ncol = 1), heights = c(1, 3, 0.5))
old_par <- par(mar = c(2.5, 1, 0.6, 0.6))

plotAverageProfile(
  testdata, pattern = "mcols",
  xlabels = c("", "middle", ""), ylim = c(0, 4)
)


plotStackedProfile(
  testdata, pattern = "mcols",
  x_labels = c("", "middle", ""),
  zlim = c(0, 3), palette = colorRampPalette(c("white", "dodgerblue", "black")),
  title = "Mark level"
)
plotStackProfileLegend(
  zlim = c(0, 3),
  palette = colorRampPalette(c("white", "dodgerblue", "black"))
)

par(old_par)
layout(1)

```

